---
title: "03-Filter_to_SI_S41"
author: "Fiona Burns/ Nick Isaac" 
opened: "23.01.26"
root.dir: "/data"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
---

This script takes the outputs from steps 1 (birds) and 2 (everything else) and converts into a cleaned, tidied dataset that is suitable for running through the multi-species indicator code. The key steps include:\ 

1) combining the datasets
2) filtering for those on 
  a) Annex 2 of the Biodiversity Regulations (aka the "SI" or the "D4a list"), and 
  b) the Priority Species list (aka the "Four country list)
3) converting the `index` column from the measurement scale to the unbounded log scale.

### load libraries 
```{r, echo = FALSE, message=FALSE}
library(reshape);library(plyr);library(dplyr);library(ggplot2);library(stringr);library(readxl); library(DescTools);
library(gridExtra);library(cowplot)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
tidydataDir <- "/data/tidy_data"
```

# LOAD DATA 

## Load species time-series

From scripts:

- 01-tidy_bird_esai26 
- 02-tidy_othergrps_esai26

```{r read_spp}
bird    <- read.csv("/data/tidy_data/intermediate_files/bird_compiled_engto24_120326.csv")
bird    <- bird %>% 
  mutate(group = "birds") %>% 
  dplyr::rename(survey = .id)

lep_mam <- read.csv("/data/tidy_data/intermediate_files/lep_mam_vp_compile_engto24_120326.csv") 
lep_mam <- lep_mam %>% 
  dplyr::rename(survey = .id) 

species_f <- list(bird, lep_mam) %>% 
  ldply()

rm(bird, lep_mam)
```


## Load Species Lists

The D4a species list and S41 list are for filtering and species dictionary to help name matching\ 

Using Four Country list generated by FourCountryList.r, "S:\\ConSci\\DptShare\\Fiona_Burns\\All_nature_projects\\RSPB_BIndicators\\Country_Lists")

Change species names to lower case for pattern matching 
```{r read_slists}
#SI plus name mismatching table
d4_sppl     <- 
  read_excel("/data/species-lists/Species List_SI_0523.xlsx", sheet = "listForSI") %>% 
  mutate(across(`Scientific name`:`Common name`, ~ str_to_lower(.x)))

name_lookup <- 
  read_excel("/data/species-lists/Species List_SI_0523.xlsx", 
             sheet = "spName_Mismatch_UKSI_Lookup") %>% 
  mutate(across(where(is.character), ~ str_to_lower(.x)))

#four country list containing the S41 species
FCL <- read.csv("/data/species-lists/FourCList_100913.csv")
n_distinct(FCL$NBN_Name[FCL$England_S41A == "Y"])
S41 <- FCL %>% 
  filter(England_S41A == "Y") %>% 
  mutate(across(where(is.character), ~ str_to_lower(.x)))
rm(FCL)

#uk species dictionary and add NBN species code to all species time-series datasets
ukspp   <- 
  read.csv("/data/species-lists/UKSI 20181220 Simplified Copy_latin.csv") %>% 
  mutate(across(where(is.character), ~ str_to_lower(.x)))

species_f$nbn_code <- 
  ukspp$NBN_TAXON_VERSION_KEY_FOR_RECOMMENDED_NAME[match(species_f$scientific_name, 
                                                         ukspp$TAXON_NAME)] 
S41$nbn_code  <- 
  ukspp$NBN_TAXON_VERSION_KEY_FOR_RECOMMENDED_NAME[match(S41$NBN_Name, ukspp$TAXON_NAME)]
```


Combine datasets

This is just a hang-over from when we had to versions of the dataset and did everything on both in a list. Retain for now, but if decide to proceed just with the freeman method, we can simplify
```{r comb_dat}
spp        <- list(species_f)
names(spp) <- c("freeman")
spp1       <- spp

rm(species_f, spp)
```


### Summarise species lists

Check the number of species in each of groups on D4 (all species) list and S41 (priority species)
```{r slist_sum}
d4_sppl %>% 
  group_by(Group) %>% 
  dplyr::summarise(no = n()) 

S41 %>% 
  group_by(Group) %>% 
  dplyr::summarise(no = n()) %>% 
  filter(Group == "bird"| Group == "insect - butterfly"| Group == "insect - moth"| Group == "terrestrial mammal"|
           Group == "flowering plant"|Group == "insect - hymenopteran"|Group == "bony fish (Actinopterygii)")
#hard to check for fw invs as cover so many taxa
```

# FILTER TO SPECIES' LISTS

## All species - d4a

### Check for mismatches

First check for mismatches between the data and the D4 (all species) list
```{r}
Uspecies       <- unique(spp1$freeman[,c("survey","group", "scientific_name", "common_name")])
Uspecies$match <- Uspecies$scientific_name %in% d4_sppl$`Scientific name`
nrow(Uspecies)
```
There are 1319 names in total. Let's figure out how many names match, and how many don't. 

```{r}
compareLists <- function(A, B){
  return(c(
    nTot = length(union(A,B)),
    nMatch = length(intersect(A,B)),
    nAnotB = length(setdiff(A,B)),
    nBnotA = length(setdiff(B, A))
  ))
}

compareLists(Uspecies$scientific_name, d4_sppl$`Scientific name`)
```
First thing to note is that there are more names (1373) than there are on the D4a list (1195) or in the dataset (1319). The number of species on the SI (all species list d4) that have not yet been matches is 54. Which taxa groups are they in?


```{r}
# Finding unmatched names
SI_names_unmatched <- setdiff(
  tolower(d4_sppl$`Scientific name`), 
  tolower(Uspecies$scientific_name)
)

# Subset d4_sppl based on case-insensitive matching
SI_not_yet_matched <- subset(d4_sppl, tolower(`Scientific name`) %in% SI_names_unmatched)

# Generate a frequency table of Survey values
with(SI_not_yet_matched, table(Survey))
```
- There are some mismatches that we're aware of from previous year. We need to check code removes: 
  - 4 moths from priority BC monitoring whose trends aren't rep of England
  - 8 birds not matched
  - 1 bee
  - 5 fish
  - 6 plants
  - 18 moths
  - 2 butterflies
  - goose subsp needed for S41 list (Anser albifrons subsp albifrons, Branta bernicla subsp. bernicla)


Here we try to match species names in three ways

- scientific name in species time-series (spp1) to scientific name in species list (d4_spp1) (column sls)
- common name in species time-series to common name in species list (column slc)
- scientific name in species time-series to scientific name in mismatch table (name_lookup) (column sllk)
A match is considered if there is a match in one or more of these, summarised in scientific_name_SL column
```{r}
spp2 <- lapply(spp1, function(x) x %>% 
                 mutate(sls = 
                          d4_sppl$`Scientific name`[match(scientific_name,
                                                          d4_sppl$`Scientific name`)], 
                          #match names by sci name
                        slc = ifelse(!is.na(common_name),
                                 d4_sppl$`Scientific name`[match(common_name, d4_sppl$`Common name`)],
                                 NA), #match names by common name, make sure catches NAs 
                        sllk = 
                          name_lookup$Name_for_StatutoryInstrument[match(scientific_name, 
                                                                         name_lookup$`Scientific name`)]) %>% 
                          #check name lookup
                 # Final step: Select best available match for scientific_name_SL
                mutate(scientific_name_SL = ifelse(!is.na(sls), sls,
                                                   ifelse(!is.na(slc), slc, 
                                                          sllk)))) 
```               


How many species are in the species time-series but have not been matched to the species list by any of the above methods?
Create a dataframe of the unmatched species to explore if correct to remove or if there are any naming mismatches (no_matchSI)
```{r}
# Count unique unmatched species names 
length(unique(
  tolower(spp2$freeman$scientific_name[is.na(spp2$freeman$scientific_name_SL)])
))

# Subset spp2$freeman where scientific_name_SL is NA 
no_matchSI <- spp2$freeman[is.na(spp2$freeman$scientific_name_SL), ]
```
Ok, there are 139 species that are sill not matching 

```{r}
compareLists(unique(tolower(spp2$freeman$scientific_name)),    tolower(d4_sppl$`Scientific name`))
compareLists(unique(tolower(spp2$freeman$scientific_name_SL)), tolower(d4_sppl$`Scientific name`))
```

Just to be clear, the `scientific_name` column has not been changed. We're matching on `scientific_name_SL`. This has created 36 extra matches.

The number of unmatched species in the dataset has apparently been brought down from 177 to 2, but in fact most are NAs.
```{r}
setdiff(unique(spp2$freeman$scientific_name_SL), d4_sppl$`Scientific name`)
```

Create a dataset with the remaining species from the SI that aren't matched.
```{r}
SI_names_unmatched2 <- setdiff(unique(d4_sppl$`Scientific name`), spp2$freeman$scientific_name_SL)
SI_still_not_yet_matched <- subset(d4_sppl, `Scientific name` %in% SI_names_unmatched2)
table(SI_still_not_yet_matched$Survey)
```
Matching the remaining name is the holy grail! This is where the work lies!


### Examine mismatches
Let's have a look at them, taxon by taxon.


#### Birds
```{r}
no_matchSI %>%
  filter(group == "birds") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```
-   Need to check if we need to keep in dark-bellied brent goose - No, only for priority 
-   Correctly removing great white egret not on SI. -These will get removed at a later point in the script.
-   Mismatch in common and latin name of shag, need to address further down
-  Also mismatch in redpoll name. acanthis flammea

#### Mammals
```{r}
no_matchSI %>%
  filter(group == "mammals") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```
- No mismatches, no action required. 

#### FW invertebrates
```{r}
no_matchSI %>%
  filter(group == "fw invertebrates") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```
- FW Inv: One mismatch to SI - Physa spp., remove for now and check if Physa fontalis, which is missing in data

```{r missingFWI}
SI_still_not_yet_matched %>% filter(Group == "MacroInvertebrates")
```
We have Physa spp. in the data and Physa frontinalis in the SI. NJBI has an email from QMUL dated 28 September 2022 confirming that a mistake was made in the dataset and that Physa spp. should be renamed as Physa fontinalis. We'll do this further down below.

The only other mismatch is Simulium. It's not entirely clear to me why it doesn't appear in the no_matchSI list. I think the issue here is that the subgenus name has been stripped out of what appears in the data. FWIW this was included in the offical D4a list because there are several other species of Simulium that are on the list: the point is that species in the subgenus Wilhemia have been lumped together, but others within Simulium are recorded at species level.

We'll fix both of these further down.


#### RIS moths
```{r}
no_matchSI %>%
  filter(survey == "ris") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```
- RIS moths: 12 new RIS moths and six missing ones 

```{r missingMoth1}
SI_still_not_yet_matched %>% 
  filter(Survey == "Rothamsted Insect Survey")
```


#### Priority moths
```{r}
no_matchSI %>%
  filter(survey == "prior_m") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```
- these six moths aren't in the SI, so should be removed. There is code to do this later, so no further action required.


#### Bumblebees
```{r}
no_matchSI %>%
  filter(group == "bumblebees") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```

```{r missingBB}
SI_still_not_yet_matched %>% 
  filter(Survey == "BeeWalks")
```

So both species are on the list. When counted as separate species this refers to the queens, which can be separated. The reason they're jointly listed is that workers cannot be distinguished, hence "lucter" = "lucorum + terrestris". There is a case for combining the index data for all three into a single composite. For now, we'll just use the data for lucter. This is handled later, so no further action at this stage.

#### Butterflies
```{r}
no_matchSI %>%
  filter(group == "butterflies") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```
- No mismatches no action required SM 10.02.26


#### Fish
```{r}
no_matchSI %>%
  filter(group == "fish") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```

```{r missingfish}
SI_still_not_yet_matched %>% 
  filter(Survey == "EA Fish")
```
- Crucian carp not in species time-series, but on SI


#### Vascular plants
```{r}
tempPlant <- no_matchSI %>%
  filter(group == "vascular plants") %>% 
  reframe(scientific_name = unique(scientific_name),
            common_name = unique(common_name))
```
-   lots of plants in dataset but not on SI

```{r missingplants}
SI_still_not_yet_matched %>% 
  filter(Survey == "NPMS")
```

Some of these clearly do match up.

- Atriplex -> Atriplex spp.
- Limonium -> Limonium spp.
- Triglochin maritima -> Triglochin maritimum

As of 2026, the England data are ready and we now have all species barring sea arrowgrass, follow up with data providers


#### Common name matching check 

Get species that only matched on common name
```{r}
cn_match <- spp2$freeman %>% 
  mutate(cname_mtch = ifelse(is.na(sls) & is.na(sllk)& !is.na(slc), "Y", "N")) %>% 
  filter(cname_mtch == "Y") %>% 
  reframe(scientific_name = unique(scientific_name),
         common_name      = unique(common_name), 
         .by = survey)

cn_match
```
- 20 species
- 4 birds, Ruff correct matching, white-fronted goose needed to remove as subsp. common name in data matches specific name in SI, lesser whitethroat - correct matching, Red grouse, 2 scientific names: lagopus scotica
lagopus lagopus. keep lagopus scotica

- 2 butterflies, correct matching
- one priority moth, Marsh mallow moth, notes in SI say only subsp in UK to on SI at species level, correct matching 
- 11 RIS moths [all subsp. in data, but specific level on SI - UKCEH to check with RIS analyst, no notes on SI list]
- 2 fish Golden grey mullet and thin lipped mullet - correct matching


### Fix mismatches

From the above code we can see there are a few naming mismatches to correct for:

- swap genera name for Chelon spp to Liza
- swap scientific name for white-tailed/buff tailed bumblebee to match that on the species list
- remove the white-fronted goose sub-species matched in error to the species list
- swap scientific name for redpoll from to Acanthis flammea to Acanthis cabaret
- swap scientific name for red grouse from lagopus lagopus to lagopus scotica
- correct plant mismatches, atriplex, limonium, triglochin

```{r}
spp3 <- lapply(spp2, function(x) x %>% 
  mutate(scientific_name_SL = ifelse(is.na(scientific_name_SL) & grepl("chelon", scientific_name) == TRUE, 
         gsub("chelon", "liza", scientific_name), #swap Chelon spp to Liza
         as.character(scientific_name_SL))) %>% 
  mutate(scientific_name_SL = 
           ifelse(scientific_name == "bombus lucter", #change B. lucter to B. lucorum/terrestris
                  "bombus lucorum/terrestris", as.character(scientific_name_SL))) %>%  
  mutate(scientific_name_SL = 
           ifelse(scientific_name == "anser albifrons subsp. albifrons", NA, 
                  scientific_name_SL)) %>% #remove goose subsp
  mutate(scientific_name_SL = ifelse(scientific_name == "physa spp.", 
                                     "physa fontinalis", scientific_name_SL)) %>%  
  mutate(scientific_name_SL = ifelse(scientific_name == "simulium spp.", 
                                     "simulium (wilhemia) spp.", scientific_name_SL)) %>%     
  mutate(scientific_name_SL = ifelse(scientific_name == "atriplex", 
                                     "atriplex spp.", scientific_name_SL)) %>%     
  mutate(scientific_name_SL = ifelse(scientific_name == "limonium", 
                                     "limonium spp.", scientific_name_SL)) %>%     
  mutate(scientific_name_SL = ifelse(scientific_name == "triglochin maritima", 
                                     "triglochin maritimum", scientific_name_SL)) %>%     
  mutate(scientific_name_SL = ifelse(scientific_name == "gulosus aristotelis", 
                                     "phalacrocorax aristotelis", scientific_name_SL)) %>% #fix shag synonym
  mutate(scientific_name_SL = ifelse(scientific_name == "acanthis flammea", 
                                       "acanthis cabaret", scientific_name_SL)) %>% #fix redpoll sci name to match species list
  mutate(scientific_name_hbw4 = ifelse(scientific_name_hbw4 == "lagopus lagopus", 
                                         "lagopus scotica", scientific_name_SL)) %>%
  mutate(sl_notes = d4_sppl$Note[match(scientific_name_SL, d4_sppl$`Scientific name`)])) #add in notes from SI
```   

Now we have fixed the naming issues how many species now match to the species or otherwise
```{r}
temp <- spp3$freeman %>% 
  mutate(matches_SI = ifelse(is.na(scientific_name_SL), "N", "Y")) 

(taxonMatch <- reshape2::dcast(unique(temp[,c("group","scientific_name", "matches_SI")]), 
                               group~matches_SI, value.var = "scientific_name", fun = function(x) length(unique(x))))
```


```{r}
sum(taxonMatch[,3])
```

So we have 1185 matches now, which is pretty close to the target of 1195.

One last time, let's look at what is missing
```{r}
compareLists(spp3$freeman$scientific_name_SL, d4_sppl$`Scientific name`)
```

Which is the one species in the dataset that doesn't match?
```{r}
setdiff(spp3$freeman$scientific_name_SL, d4_sppl$`Scientific name`)
```

Ah - it's NA!

And a final look at those 10 species that will not be in the index this time
```{r}
MissingFromIndex <- setdiff(
  tolower(d4_sppl$`Scientific name`), 
  tolower(spp3$freeman$scientific_name_SL)
)

MissingFromIndex <- setdiff(d4_sppl$`Scientific name`, spp3$freeman$scientific_name_SL)
subset(d4_sppl, tolower(`Scientific name`) %in% MissingFromIndex)
```

Potential development option: Investigate species on SI with no data at present

-   The six moths are ones that the RIS analyst warned me about: they are species that failed the automated checks. So we need to revisit these at a later date.
-   The Crucian Carp is a bit of an oddity: there was some discussion about whether it should be included and it's possible that QMUL excluded both this and the Common Carp. The latter nearly made it onto the SI but was excluded as it is an invasive species.
-   We no longer have data for sea arrowgrass - follow-up with data provider Triglochin maritima - expect this to be available in future years
-   We no longer have data for Scythris siccella following a review of data robustness by BC
-   We only have six year of data for Straw Belle following the same review, should be able to reintroduce in a few years

### Remove mismatches
 
Now we have fixed naming issues we will remove non-matching species from the dataset and tidy it up 
Use the name on the species list as 'scientific_name' and the name in the species time-series as 'scientific_name_data'
```{r}    
spp_d4 <- lapply(spp3, function(x) x %>%     
  filter(!is.na(scientific_name_SL)) %>%  # Remove unmatched species
  mutate(
    scientific_name_data = scientific_name) %>%  # Save sci name used in dataset
  select(group, survey:common_name, scientific_name_data, scientific_name_SL, 
         nbn_code, year, index, se, sd_unsm, lcl_unsm,ucl_unsm, index_orig, index_sm:ucl_sm, se_sm, index.raw, sl_notes) %>% 
  rename(scientific_name = scientific_name_SL))

# Check number of species in filtered dataset by group
spp_d4$freeman %>% 
  group_by(group) %>% 
  summarise(n = n_distinct(scientific_name_data)) 

rm(spp2)
```

### Output match to SI

Create a version of the SI list showing which species we have data for this year or otherwise\ 
Check how many species we are missing data for by taxonomic group and save file
```{r}
d4_sppla <- d4_sppl %>% 
  mutate(data = match(tolower(`Scientific name`), tolower(spp_d4$freeman$scientific_name))) 

# How many species are we missing data for?
d4_sppla %>% 
  mutate(data_bin = ifelse(is.na(data), "N", "Y")) %>% 
  reframe(.by = c(Group, data_bin), n = n_distinct(tolower(`Scientific name`))) 
```

save match to SI
```{r, eval=FALSE}
write.csv(d4_sppla, file = file.path(tidydataDir, "SIlist_data_notdata_datato2024_120326.csv"), row.names = FALSE)
```



## Priority species - d6a

### Check for mismatches

* We know that willow tit and lesser spotted woodpecker are in the data at a species level but on the S41 as the only subsp present in UK. Switch to subsp name
* Try to match between the species time-series (spp1) and the S41 list based on 
    - matching scientific names (pr_s)
    - matching common names (pr_c)
    - matching nbn code (pr_n)
* Assume a match if there is a match in any one of the above (pr_e)

```{r}
spp_d6 <- spp1
spp_d6$freeman <- spp_d6$freeman %>%                    
  mutate(
    scientific_name = ifelse(tolower(scientific_name) == tolower("poecile palustris"), 
      "poecile palustris subsp. palustris/dresseri",
      ifelse(tolower(scientific_name) == tolower("dendrocopos minor"), 
        "dendrocopos minor subsp. comminutus", 
        scientific_name))) %>% 
  mutate(
    pr_s = S41$NBN_Name[match(scientific_name, S41$NBN_Name)], #matching on sci name
    pr_n = ifelse(is.na(nbn_code), NA, 
                  S41$NBN_Name[match(nbn_code, S41$nbn_code)]), #matching on nbncode
    pr_c = ifelse(is.na(common_name), NA, 
                  S41$NBN_Name[match(common_name, S41$Common_Name)] #matching on common name
    )) %>% 
  mutate(pr_e = ifelse(!is.na(pr_s), pr_s,
                        ifelse(!is.na(pr_c), pr_c, pr_n)))
```
 
Let's see what has matched by group in cf. to total number on S41 list
```{r}
spp_d6$freeman %>% 
  filter(!is.na(pr_e)) %>% 
  group_by(group) %>% 
  dplyr::summarise(n_spp = n_distinct(scientific_name))

S41 %>% 
  group_by(Group) %>% 
  dplyr::summarise(no = n()) %>% 
  filter(Group == "bird"| Group == "insect - butterfly"| Group == "insect - moth"| Group == "terrestrial mammal"|
           Group == "flowering plant"|Group == "insect - hymenopteran"|Group == "bony fish (Actinopterygii)")
```

So we have 

* 43 birds out of 49 
* 21 butterflies out of 23 
* 77 moths out of 142 
* 8 terrestrial mammals out of 17
* 1 bumblebee - Bombus humilis
* 2 plants - Juniper and one mistaken mismatch -  Carex otrubae/ aka Carex vulpina (which is on the S41!)
* 2 fw invertebrates - white clawed crawfish and yellow mayfly
* 9 fish

### Examine mismatches

#### Birds

Are we happy with the bird selection?

-   Which species are we missing?
-   Which species are in the data but not matched?
```{r}
#priority species with data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(!is.na(in_data),
         Group == "bird") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
#priority species without data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(is.na(in_data),
         Group == "bird") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
```
* Six species not matched, we don't have robust data for 5 of  these species so happy with the matching
* ACTION REQUIRED: Need to match redpoll - in data listed as acanthis flammea, s41 list as carduelis cabaret

#### Butterflies

Are we happy with the butterfly selection
```{r}
#priority species with data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(!is.na(in_data),
         Group == "insect - butterfly") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
#priority species without data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(is.na(in_data),
         Group == "insect - butterfly") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
```
* Only 2 species without data are the Mountain Ringlet, we don't have reliable data for this species and Large Heath, data variably available in UKBMS. Suggest we don't include if don't have a reliable trend every year. 

No action required.

#### Moths

Are we happy with the moth selection?
```{r}
#priority species with data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(!is.na(in_data),
         Group == "insect - moth") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
#priority species without data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(is.na(in_data),
         Group == "insect - moth") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
```
* Dark Bordered Beauty and Silky Wave priority moth data not suitable for England 
* Bright Wave Moth on S41 as subsp. (Idaea ochrata subsp. cantiata) data for species, but appropriate to use as only subspecies in UK
* Straw Belle on S41 as subsp. (Aspitates gilvaria subsp. gilvaria) data for species, but appropriate to use as only subspecies in UK
* As described in line 270, there are six fewer species on the RIS dataset this year. Four of these are priority species, so we have four fewer species compared to the previous iteration of D6a (Crescent, Grey Mountain Carpet, Haworth's Minor, Heath Rustic)
* ACTION REQUIRED: Remove matching for DBB and Silky Wave, add match for Straw Belle and Bright Wave Moth [NB we don't have straw belle data now]

#### Mammals

Are we happy with the mammal selection?
```{r}
#priority species with data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(!is.na(in_data),
         Group == "terrestrial mammal") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
#priority species without data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(is.na(in_data),
         Group == "terrestrial mammal") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
```
* 9 mammals not matching - We don't have data for them
- the eight matches are correct

No action required

#### Vascular plants

Are we happy with the vplant selection?
```{r}
#priority species with data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(!is.na(in_data),
         Group == "flowering plant"|Group == "conifer") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(#Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
#priority species without data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(is.na(in_data),
         Group == "flowering plant"|Group == "conifer") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(#Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
```
- carex vulpina (mismatch with carex otrubae) pr_e 'carex vulpina', remove later in script
- juniper correct

NB 100326 checked the new plant species with data this year against synonyms in NBN Atlas and couldn't find any additional matches

ACTION REQUIRED: Remove matching for Carex otrubae

#### Fish

Are we happy with the fish selection?
```{r}
#priority species with data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(!is.na(in_data),
         Group == "bony fish (actinopterygii)") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
#priority species without data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(is.na(in_data),
         Group == "bony fish (actinopterygii)") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
```

Eight species with data
lesser sandeel, eel, herring, cobitis taenia, smelt, plaice, atlantic salmon, brown trout

No action required

#### Freshwater invertebrates

Are we happy with the fw inv selection? 

-   not sure how to select here, lots of different groups covered
```{r}
#priority species with data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(!is.na(in_data),
         Group == "crustacean"| Group == "mollusc"|
           grepl("insect", Group) == TRUE) %>% 
   filter(Group != "insect - moth",
         Group != "insect - butterfly",
         Group != "insect - hymenopteran") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Group,
                   Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
#priority species without data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(is.na(in_data),
         Group == "crustacean"| Group == "mollusc"|
           grepl("insect", Group) == TRUE) %>%
  filter(Group != "insect - moth",
         Group != "insect - butterfly",
         Group != "insect - hymenopteran") %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Group,
                   Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
```
-   two species, white-clawed crayfish, yellow mayfly - out of 153 [some won't be freshwater]

No action required


#### Bumblebees

Are we happy with the bbee selection?
```{r}
#priority species with data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(!is.na(in_data),
         grepl("bombus", NBN_Name) == TRUE) %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   NBN_Name = unique(NBN_Name))
#priority species without data
S41 %>% 
  mutate(in_data = match(NBN_Name, spp_d6$freeman$pr_e)) %>% 
  filter(is.na(in_data),
        grepl("bombus", NBN_Name) == TRUE) %>% 
  group_by(NBN_Name) %>% 
  dplyr::summarise(Common_Name = unique(Common_Name),
                   DesigName = unique(DesigName))
```
Single species Bombus humilis - Brown-banded carder-bee

No action required


Summary of corrections needed for naming mismatches

* Carex otrubae incorrectly labelled in UK species dictionary and therefore incorrectly flagged as on the S41, remove match
* Bright Wave and Straw Belle incorrectly not matched (lines 597:594)
* Silky Wave and Dark Bordered Beauty not good England data, remove match
* Remove Large Heath match, as data availability intermittent
* Two species called grayling, a butterfly and a fish, only the butterfly is on the S41
* Need to match redpoll - in data listed as acanthis flammea, s41 list as carduelis cabaret
* Summarise remaining mismatches and matches


### Fix mismatches

```{r correct mismatches}
spp_d6a <- lapply(spp_d6, function(x) x %>% 
                   mutate(
      scientific_name = ifelse(scientific_name == "acanthis flammea",
                               "carduelis cabaret",
                               scientific_name),
      pr_e = 
        ifelse(scientific_name == "carduelis cabaret", "carduelis cabaret",
                    pr_e)) %>% 
                   mutate(pr_e = ifelse(scientific_name == "carex otrubae"| 
                                          scientific_name == "thymallus thymallus",
                                        NA, as.character(pr_e))) %>% 
                   mutate(pr_e = ifelse(is.na(common_name), as.character(pr_e),
                                     ifelse(common_name == "silky wave"|
                                              common_name == "dark bordered beauty"|
                                            common_name == "large heath", NA,
                                         ifelse(common_name == "bright wave moth", 
                                                "aspitates gilvaria subsp. gilvaria",
                                        as.character(pr_e))))))
```


### Remove mismatches

Filter to S41 list by selecting those species matched and tidy up columns
```{r filter to s41}
spp_d6b <- lapply(spp_d6a, function(x) x %>% 
                    filter(!is.na(pr_e))   %>% 
                    select(-uk_spp, -(pr_s:pr_c), -best_d) %>% 
                    select(group, survey:scientific_name, nbn_code, year, index,
                           index_orig, se, sd_unsm, lcl_unsm, ucl_unsm, index_sm:ucl_sm, se_sm,index.raw, pr_e))

ddply(spp_d6b$freeman, .(group), summarise, species = n_distinct(scientific_name))
```
-   same species as last year
-   barring clumping of common and lesser redpoll into a single species

### Output match to S41 List

Version of S41 list showing which species we have data for
```{r}
S41_data <- S41 %>% 
   mutate(in_data = ifelse(is.na(match(NBN_Name, spp_d6b$freeman$pr_e)), "N", "Y")) %>%
   select(NBN_Name, nbn_code, Common_Name, HGroup, Group, in_data)

ddply(S41_data, .(HGroup, Group, in_data), summarise, species = n_distinct(NBN_Name))
```

save match to S41
```{r, eval=FALSE}
write.csv(S41_data, file = file.path(tidydataDir, "S41list_data_notdata_120326.csv"), row.names = FALSE)
```


# START YEAR PER GROUP

Which year to start from? 
```{r}
spp_c_min <- function(data) {ddply(data, .(scientific_name), dplyr::summarise,
               min_yr =  min(year[!is.na(index)], na.rm = TRUE), 
               max_yr =  max(year[!is.na(index)], na.rm = TRUE),
               group = unique(group))
}
spp_m  <- lapply(list(spp_d4$freeman, spp_d6b$freeman), spp_c_min) 

lapply(spp_m, function(x) table(x$min_yr))
lapply(spp_m, function(x) table(x$max_yr))
lapply(spp_m, function(x) table(x$min_yr, x$group))
```
* start overall in 1970
* when to start group metrics?  moths 1970, birds 1970, butterflies 1976 mammals 1995 (98 for D6a), vp 2015, fw_i = 2013, fish 2000 


### Truncate species time-series  

* remove all data for years prior to 1970 and for mammals prior to 1995 (D6a prior to 98) 
```{r}
spp_d4a <- lapply(spp_d4, function(x) x %>% 
  filter(year > 1969) %>% 
  filter(!(group == "mammals" & year < 1995)))

spp_d6c <- lapply(spp_d6b, function(x) x %>% 
  filter(year > 1969 ) %>% 
  filter(!(group == "mammals" & year < 1998)))
```

# OUTPUT FILES

### Create species list per Indicator
```{r spp_list, eval=FALSE}
spp_list_d4a <- spp_d4a$freeman %>% 
  group_by(scientific_name, scientific_name_data, common_name) %>% 
  summarise(group = unique(group),
            survey = unique(survey))

spp_list_d6c <- spp_d6c$freeman %>% 
  group_by(scientific_name, common_name) %>% 
  dplyr::summarise(group = unique(group),
            survey = unique(survey))
```

save species lists
```{r, eval=FALSE}
write.csv(spp_list_d4a, file = file.path(tidydataDir, "spp_in_d4a_to_2024_120326.csv"), row.names = FALSE)
write.csv(spp_list_d6c, file = file.path(tidydataDir, "spp_in_d6a_to_2024_120326.csv"), row.names = FALSE)
```

### Save species time-series

- put index on log scale
- save measurement scale index as 'index_m'
NB prior to March 2024, the `index` column for plants was already on the log scale. Now it's treated the same as all other datasets.
```{r}
spp_d4b <- lapply(spp_d4a, function(x) x %>% 
  dplyr::mutate(index_l = log(index)) %>% 
  dplyr::rename(index_m = index, species = scientific_name) %>% 
  dplyr::rename(index = index_l) %>% 
  select(group:year, index, index_m:sl_notes) %>% #this keeps all columns, just puts index near the front
  arrange(group, species, year)) 

spp_d6d <- lapply(spp_d6c, function(x) x %>% 
  dplyr::mutate(index_l = log(index)) %>% 
  dplyr::rename(index_m = index, species = scientific_name) %>% 
  dplyr::rename(index = index_l) %>% 
  arrange(group, species, year) %>% 
  select(group:year, index, index_m:pr_e))
```

save two sets of species time-series 
```{r write_timeseries, eval=FALSE}
write.csv(spp_d4b$freeman, file = "/data/tidy_data/d4_spp_timeseries_engto24_120326.csv", row.names = FALSE)
write.csv(spp_d6d$freeman, file = "/data/tidy_data/d6_spp_timeseries_engto24_120326.csv", row.names = FALSE)
```
