---
title:  "04 - metadata_d4a_d6a_to22"
author: "Fiona Burns"
opened: "23.01.26"
root.dir: "/data"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
---

 - This script takes the output of '03-Filter to SI_S41' script and produces some summaries.
 - It does not change the output files, which are further used in script '05-preSmoothing'

```{r, include = FALSE, echo=FALSE}
library(reshape);library(plyr);library(dplyr);library(ggplot2);
library(readxl); library(DescTools);
library(gridExtra);library(cowplot)
```

```{r}
source("/data/notebooks/rstudio-esaiukpsi26/tidy_functions.R")
```

set up folder paths
```{r setup}
tidydataDir <- "/data/tidy_data" 
figDir_d4   <- "/data/notebooks/rstudio-esaiukpsi26/england/species_d4a/figures"
figDir_d6   <- "/data/notebooks/rstudio-esaiukpsi26/england/priority_d6a/figures"
```

read time-series
```{r read ts}
spp_d4a <- list(read.csv(file.path(tidydataDir,"d4_spp_timeseries_engto24_120326.csv")))
names(spp_d4a) <- c("freeman") 

spp_d6a <- list(read.csv(file.path(tidydataDir,"d6_spp_timeseries_engto24_120326.csv")))
names(spp_d6a) <- c("freeman") 
```


## Metadata 

### All species - D4a

#### Total number of species 

 - the number of species, 
 - the number of species per group and
 - the number of species per survey per group
```{r meta d4}
meta_d4   <- spp_d4a$freeman %>% 
   summarise(nspp = n_distinct(species), 
      min_yr = min(year[!is.na(index)])[1],  
      max_yr = max(year[!is.na(index)])[1]) %>% 
  mutate(group = "ALL")

meta_d4_g <- spp_d4a$freeman %>% 
     group_by( group) %>% 
    summarise(nspp = n_distinct(species), 
      min_yr = min(year[!is.na(index)])[1],  
      max_yr = max(year[!is.na(index)])[1]) %>% 
    ungroup() %>% 
  full_join(meta_d4)

meta_d41 <- spp_d4a$freeman %>% 
     group_by(survey, group) %>% 
     summarise(nspp = n_distinct(species), 
      min_yr = min(year[!is.na(index)])[1], 
      max_yr = max(year[!is.na(index)])[1]) %>% 
    ungroup() %>% 
    full_join(meta_d4_g) %>% 
    arrange(group, desc(nspp)) %>% 
    select(group, survey, nspp, min_yr, max_yr ) %>% 
    mutate(survey = ifelse(is.na(survey), "TOTAL", as.character(survey)))
meta_d41
```

```{r, eval=FALSE}
write.csv(meta_d41, 
          file = file.path(tidydataDir, "metadata_d4engto24_120326.csv"), row.names = FALSE)
```


#### Number of species per year

Let's look at the number and proportion of species with values in the dataset in each year

 - summarise across all species
 - summarise by taxonomic group
 - add in total number of species from meta... file above
 - express all in relation to the total species
 
```{r spp yr d4}
sppyr_d4a <- spp_d4a$freeman %>% 
  group_by(year) %>% 
  summarise(spp_yr = n_distinct(species[!is.na(index)]) ) %>% 
  mutate(group = "ALL") 
sppyr_grp_d4a <- spp_d4a$freeman %>% 
  group_by(year, group) %>% 
  summarise(spp_yr = n_distinct(species[!is.na(index)])) %>% 
  full_join(sppyr_d4a) %>% 
  mutate(total_spp = 
           meta_d4_g$nspp[match(group, meta_d4_g$group)]) %>% 
  mutate(pspp = spp_yr/total_spp * 100)
```

```{r, eval=FALSE}
write.csv(sppyr_grp_d4a, file = file.path(tidydataDir, "spp_in_engd4ato24_peryr_120326.csv"))
```

plot
```{r}
p <- ggplot(sppyr_grp_d4a, aes(year, spp_yr)) +
  geom_col()+
  facet_wrap(vars(group), scales = "free_y")+
  ylab("Number of species")
p
```

```{r, eval=FALSE}
jpeg(file = file.path(figDir_d4, "d4_engto24_spp_in_ind_grp.jpeg"),
     width = 7, height = 5, units = "in", res = 500)

print(p)
dev.off()
```

### Priority species - D6a

#### Total number of species 

- the number of species, 
- the number of species per group and
- the number of species per survey per group
```{r d6 meta}
meta_d6   <- spp_d6a$freeman %>% 
   dplyr::summarise(nspp = n_distinct(species), 
      min_yr = min(year[!is.na(index)])[1],  
      max_yr = max(year[!is.na(index)])[1]) %>% 
  mutate(group = "ALL")

meta_d6_g <- spp_d6a$freeman %>% 
     group_by(group) %>% 
    dplyr::summarise(nspp = n_distinct(species),  
      min_yr = min(year[!is.na(index)])[1], 
      max_yr = max(year[!is.na(index)])[1]) %>% 
    ungroup() %>% 
  full_join(meta_d6)

meta_d61 <- spp_d6a$freeman %>% 
     group_by(survey, group) %>% 
    dplyr::summarise(nspp = n_distinct(species), 
      min_yr = min(year[!is.na(index)])[1], 
      max_yr = max(year[!is.na(index)])[1]) %>% 
    ungroup() %>% 
    full_join(meta_d6_g) %>% 
    arrange(group, desc(nspp)) %>% 
    select(group, survey, nspp, min_yr, max_yr ) %>% 
    mutate(survey = ifelse(is.na(survey), "TOTAL", as.character(survey)))
meta_d61
```

```{r, eval=FALSE}
write.csv(meta_d61,    file = file.path(tidydataDir, "metadata_d6engto24_120326.csv"), row.names = FALSE)
```

#### Number of species per year

Let's look at the number and proportion of species with values in the dataset in each year

 - summarise across all species
 - summarise by taxonomic group
 - add in total number of species from meta... file above
 - express all in relation to the total species
```{r spp yr d6}
sppyr_d6a <- spp_d6a$freeman %>% 
  group_by(year) %>% 
  dplyr::summarise(spp_yr = n_distinct(species[!is.na(index)]) ) %>% 
  mutate(group = "ALL") 
sppyr_grp_d6a <- spp_d6a$freeman %>% 
  group_by(year, group) %>% 
  dplyr::summarise(spp_yr = n_distinct(species[!is.na(index)])) %>% 
  full_join(sppyr_d6a) %>% 
  mutate(total_spp = 
           meta_d6_g$nspp[match(group, meta_d6_g$group)]) %>% 
  mutate(pspp = spp_yr/total_spp * 100)
```

save  
```{r,eval=FALSE}
write.csv(sppyr_grp_d6a, file = file.path(tidydataDir, "spp_in_engd6ato24_peryr_120326.csv"))
```

plot
```{r}
p <- ggplot(sppyr_grp_d6a, aes(year, spp_yr)) +
  geom_col()+
  facet_wrap(vars(group), scales = "free_y")+ 
  ylab("Number of species")
p
```

```{r eval=FALSE}
jpeg(7, 5, units = "in", file  = file.path(figDir_d6, "d6_engto24_spp_in_ind_grp.jpeg"), res = 500)
print(p)
dev.off()
```

## Extreme values 

Let's look at the number of extreme values 

 - save info on how often species time-series go below at 0.01 or above 100  
 - in order to do this we first rescale the time-series to start at 1 in their first year
 - we don't need to rescale to use the freeman method, this is just to assess extreme values and is not saved into the data used to run the indicator models on

function to count the number of extreme values
```{r num cap fun}
no_cap <- function(data) {
  a_b <-  dplyr::summarise(data, 
                    above100 = sum(index_m > 100, na.rm = TRUE),
                    below0.01 = sum(index_m < 0.01, na.rm = TRUE))
return(a_b)
} 
```

##### Function now in 'tidying_functions.R'

- function to rescale species' time-series relative to the value in the first year [rescale]

Estimate the number of rapidly increasing or decreasing species
```{r num extremes}
ext_trd_d4a  <- spp_d4a$freeman %>% 
    rename(scientific_name = species) %>% 
    rescale(ind_col = index_m, Index_root = 1) %>% 
    group_by(group, scientific_name, common_name) %>% 
    no_cap() %>% 
    filter(!(above100 == 0 & below0.01 ==0))

ext_trd_d6a <- spp_d6a$freeman %>% 
  rename(scientific_name = species) %>% 
    rescale(ind_col = index_m, Index_root = 1) %>%
    group_by(group, scientific_name, common_name) %>% 
    no_cap() %>% 
    filter(!(above100 == 0 & below0.01 ==0))
```

```{r, eval=FALSE}
write.csv(ext_trd_d4a, 
          file = file.path(tidydataDir, "d4a_engto24_spp_a100_b0.001_120326.csv"), 
          row.names = FALSE)

write.csv(ext_trd_d6a, 
          file = file.path(tidydataDir, "d6a_engto24_spp_a100_b0.001_120326.csv"), 
          row.names = FALSE)
```

## Number of zeros or psuedo-zeroes

Zeros are difficult to deal with in indicator models and we use a rule of thumb workaround to address this. Good to assess:

- the number of species affected 
- the number of instances 

```{r num zeros}
zrs_d4 <- spp_d4a$freeman %>% 
  group_by(survey, species, common_name) %>% 
  summarise(zero_psuedozero = sum(index_orig == 0, na.rm = TRUE)) %>% 
  filter(zero_psuedozero != 0) 

zrs_d6 <- spp_d6a$freeman %>% 
  dplyr::group_by(survey, species, common_name) %>% 
  dplyr::summarise(zero_psuedozero = sum(index_orig == 0, na.rm = TRUE)) %>% 
  filter(zero_psuedozero != 0)
```

```{r, eval=FALSE}
write.csv(zrs_d4, 
          file.path(tidydataDir, "d4_engto24_zero_psuedozeros_inspptrends_120326.csv"), 
          row.names = FALSE)

write.csv(zrs_d6, 
          file.path(tidydataDir, "d6_engto24_zero_psuedozeros_inspptrends_120326.csv"), 
          row.names = FALSE)
```



