---
title: "run Freeman By Group"
author: "Nick Isaac"
date: "2024-02-28"
output: 
  html_document:
    toc: false
editor_options: 
  markdown: 
    wrap: 72
---

We're going to run the Freeman method for multiple groups with differing
levels of smoothing.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(devtools)
install_github('biologicalrecordscentre/BRCindicators')
library(BRCindicators)
library(jagsUI)
library(ggplot2)
library(reshape2)
library(dplyr)
##need to type 'sudo apt install jags' in terminal 
```

## Load in the data.

```{r readData}
datadir <- "/data/tidy_data" 

#to use unsmoothed VP time-series
data <- read.csv(file = file.path(datadir,"d4a_spp_ts_engto24_Smoothed_unsmVP_120326.csv")) 
```

 
Legacy code from 2023 when we compared smoothing on log and measurement scale, smoothing now all done on log scale

Specify whether to use smoothing on the log scale
(`index2`) or on the measurement scale (`index1`).
```{r}
logSmoothing <- TRUE
if(logSmoothing){
  data$index <- data$index2
} else {
  data$index <- data$index1
}
```

We need to add a small fudge factor to the fish data, because there are
no fish data for 2020. If we don't do this, the model will incorrectly
believe that 2022 is actually 2021. Sprattus sprattus
```{r}
data <- rbind(data[,c("species", "year", "group", "index")], 
              data.frame(species = "sprattus sprattus", 
                               year = 2020,
                               group = "fish",
                               index = NA))
```

## Define a function to run the model 

```{r runFreeman, echo=TRUE}
runFreeman <- function(gr, ypk, save = TRUE, runShort = TRUE, nit=5000, spTrends = FALSE, rebase = NULL){
                       
                       if(gr == "all") {
                         inddat <- data[,c("species","year", "index")]
                       } else {
                         inddat <- subset(data, group==gr, select = c("species","year", "index"))
                       }

                       n.yr    <- length(unique(inddat$year))
                       n.knots <- round(n.yr / ypk)
                       print(paste0(gr, ": ", ypk, " years per knot. With ",n.yr, " years of data = ", n.knots," knots overall"))

                       if(runShort & n.knots == 1) {
                         n.knots <- 2
                         print("running with two knots")
                       }
                       
                       if(n.knots > 1){
                   
                         mod <- bma(data = inddat, 
                                 num.knots = n.knots,
                                 parallel = TRUE,
                                 plot = FALSE,
                                 n.iter = nit,
                                 save.sppars = spTrends)
                       
                         md <- data.frame(group = gr,
                                  ypk = ypk,
                                  n.knots = n.knots,
                                  n.it = nit)
                         attr(mod, "metadata") <- md

                        if(save){
                         letter <- ifelse(logSmoothing, "L","M")
                            saveRDS(mod, file=paste0("species_d4a/results/", gr, "_",
                                                 ypk,"ypk_",letter,"_",
                                                 rebase, "_",
                                                  nit/1000,"kit_",
                                                  format(Sys.Date(),"%y%m%d"),".rds"))
                         return(NULL)
                       } else {
                         return(mod)
                       }
                       } else {
                         print("only one knot: skipping")
                       }
                       
}
```

Define some simple parameters for the model runs.
```{r setPars}
taxa <- unique(data$group)
# set the number of years per knot
ypk <- c(10, 3) 
```


## Run models

### Full time-period

#### Group level

Run this for all the groups except moths and fish
```{r, eval=FALSE}
p <- expand.grid(ypk = ypk, gr = setdiff(taxa, c("moths", "fish")))
system.time({
    output_group <- mapply(ypk = p$ypk, gr = p$gr, runFreeman, SIMPLIFY = FALSE, save = TRUE)
})
```
Runtime varies with number of species and length of time-series:

- Mammals all four in a minute 
- Fish & plants take about one minute per run  
- FW invertebrates & butterflies about 2 minutes each 
- Birds about 15 mins each So only about 75 mins to run the lot

Now for the moths

```{r,  eval=FALSE}
system.time({
    output_group <- sapply(ypk, runFreeman, gr = "moths", save = TRUE)
})
```
Each run of the moths takes just over 30 minutes

And for the fish
```{r, eval=FALSE}
system.time({
    output_group <- sapply(ypk, runFreeman, gr = "fish", save = TRUE)
})
```


#### All species

Now the all the species MSI. For this we want to save the species
trends.
```{r,  eval=FALSE}
system.time({
    output_group <- sapply(ypk, runFreeman, gr = "all", save = TRUE, spTrends = TRUE)
})
```
Just over an hour each (a bit slower if saving species parameters)


#### 1 year per knot

Finally, we'll rerun each group with one knot for every year of data.
```{r, eval=FALSE}
system.time({
    output_group <- sapply(c(taxa, "all"), runFreeman, ypk = 1, save = TRUE, spTrends = FALSE)
})
```
2 hours total

all species didn't run - do separately
```{r, eval=FALSE}
system.time({
    output_group <- runFreeman("all", ypk = 1, save = TRUE, spTrends = FALSE)
})
```

### Short-term assessment

#### Ten year assessment

- Re-run all species for last ten years with curtailed dataset [ten growth rate estimates, 11 years]
- Using 3ypk
- What about species with less than ten years span of data?
- Keep in for now, but say which these are? 

```{r}
data <- data %>% 
  filter(year > 2013)

sparse <- data %>% 
  group_by(group, species) %>% 
  summarise(yrs = max(year)-min(year)) %>% 
  filter(yrs < 10)
n_distinct(data$species)
sparse %>% arrange(yrs)
```
- 1184 species (lost Nightjar)
- Peregrine has only a single datapoint, need to remove
- all the plants and Water vole only have ten years 
- RBBP/ WeBS and some fish missing 2024

remove species with only a single data point
```{r}
data <- data %>% 
  filter(species != "falco peregrinus")
n_distinct(data$species) == 1183
```

```{r,  eval=FALSE}
system.time({
    output_group <- runFreeman(gr = "all", 
                           save = TRUE, ypk = 3, spTrends = FALSE, rebase = "d4_ten_yr")
})
```

#### Five year assessment 

- Re-run all species for last five years with curtailed dataset [five growth rate estimates, 6 years]
- Using 3ypk
```{r}
data <- data %>% 
  filter(year > 2018)

sparse_five <-data %>% 
  group_by(group, species) %>% 
  summarise(yrs = max(year)-min(year)) %>% 
  filter(yrs < 5)
n_distinct(data$species) #1182
sparse_five %>% arrange(yrs)
```
- Circus cyaneus (Hen harrier) and Lesser sandeel (Ammodytes tobianus) have one datapoint, need to remove
- Eustroma reticulata (moth) has dropped out of the dataset
- Athetis pallustris span two years 
- Pyropteron chrysidiformis spans four years
- 46 species (RBBP, WeBS, some fish) are missing 2024

remove species with a single data point
```{r}
data <- data %>% 
  filter(!species %in% c("circus cyaneus", "ammodytes tobianus"))
n_distinct(data$species) == 1180
```

re-run for last five years
```{r,  eval=FALSE}
system.time({
    output_group <- runFreeman(gr = "all", 
                           save = TRUE,ypk = 3 ,  spTrends = TRUE, rebase = "d4_five_yr")
})
```





