---
title: "Explore English abundance index to2024"
author: "Nick Isaac"
date: "2024-03-22"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
---

We have run the Freeman method for multiple groups, with differing
levels of smoothing, on a dataset of pre-smoothed index values (on the
log scale).

The underlying dataset contains 1185 taxa (mostly species), i.e. \>99%
of those listed on Schedule 2 of the Biodiversity Regulations. The 10
missing species are: - 1 plant and nine species whose model outputs
failed the automatic checks (eight moths and one fish)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(coda)
library(ggplot2)
library(reshape2)
library(plyr)
library(dplyr)
library(tidyr)
library(scales)
library(devtools)
install_github('biologicalrecordscentre/BRCindicators')
library(BRCindicators)
```

```{r}
memory.limit(size = 12000)
```

# Read in the model files

```{r readData, include = FALSE}
modeldir <- "/data/notebooks/rstudio-esaiukpsi26/england/species_d4a/results" 
qa_dir   <- "/data/notebooks/rstudio-esaiukpsi26/england/species_d4a/results/qa"
```

## Get the models

```{r setPars}
modelfiles <- list.files(modeldir, recursive = TRUE, patt = ".rds")
modelfiles <- modelfiles[(grepl("0313", modelfiles)==TRUE)] #select final run
modelfiles
```

```{r readModels}
output <- lapply(modelfiles, function(modfilename){
 readRDS(file.path(modeldir, modfilename)) 
})
```

## Get the metadata

Extract the metadata from all the model runs into a dataframe

```{r meta}
pars <- lapply(output, function(mod){
  md <- attr(mod, "metadata")
})
pars <- do.call(rbind, pars)

names(output) <- modelfiles #five_yr is model 4 and ten year is model 5, need to differentiate from all species models for full time period

pars <- pars %>% 
  mutate(group =  ifelse(grepl("five_yr", modelfiles) == TRUE, paste(group, "five_yr", sep = "_"),
                                     ifelse(grepl("ten_yr", modelfiles) == TRUE, paste(group, "ten_yr", sep = "_"),
                                            group))) 
```

## Explore and combine

Has it converged?

```{r diag}
MCMCStats <- lapply(output, function(msi) {
  dat <- attr(msi, "model")$summary
  dat[!grepl("^M", dimnames(dat)[[1]]), 8] #column 8 is rhat, column 9 is effective sample size
})
pars$Rhat3q <- sapply(MCMCStats, quantile, 0.75, na.rm=TRUE)
```

#### Extra exploration of model performance

```{r diag_theta_dev}
diagnostics  <- ldply(output, function(msi) {
  dat  <- attr(msi, "model")$summary
  rdat <- data.frame(theta_Rh     = dat[grepl("theta", dimnames(dat)[[1]]), "Rhat"],
                     theta_Nef    = dat[grepl("theta", dimnames(dat)[[1]]), "n.eff"],
                     deviance_Rh  = dat[grepl("dev", dimnames(dat)[[1]]), "Rhat"],
                     deviance_Nef = dat[grepl("dev", dimnames(dat)[[1]]), "n.eff"])
  })  %>% 
  cbind(modelfiles) %>% 
  select(5:1)
```

```{r, eval=FALSE}
write.csv(diagnostics, 
          file = file.path(qa_dir, "d4a_model_diagnostics_2026_130326.csv"), 
          row.names = FALSE)
```

Stick them all together

```{r comb}
df <- do.call(rbind, output)
attr(df, "model") <- NULL
attr(df, "metadata") <- NULL
```

Now coerce the output into something useful:

```{r Manipulate}
pars$nyr <- c(sapply(output, nrow))

df$group <- melt(mapply(rep, pars$group, pars$nyr))$value
df$ypk <- melt(mapply(rep, pars$ypk, pars$nyr))$value
df$n.knots <- melt(mapply(rep, pars$n.knots, pars$nyr))$value
df$ypk <- factor(df$ypk)
```

```{r, eval=FALSE}
write.csv(df,   file = file.path(modeldir, "MSI|group+_presmooth_allspp_to2024_130326.csv"))
write.csv(pars, file = file.path(modeldir, "SmoothingParameters_allspp_to2024_130326.csv"))
```

# Unsmoothed trends

An easier way to get the unsmoothed line for each group is to use the
WBI method for creating an MSI. Let's see if this is a sensible way to
plot unsmoothed trends. For this we need to get a version of the dataset
with interpolated values

We didn't save an interpolated version of the species time-series, so we
need to generate this in next code block. Also, this is not truly an
'unsmoothed' line, as it is using pre-smoothed time-series.

Read in the script of tidying and indicator functions

```{r src}
source("/data/notebooks/rstudio-esaiukpsi26/tidy_functions.R")
```

Read in species time-series and process so can run IndFun  This
involves:

-   Converting to measurement scale
-   Expanding the grid so that all spp/yr combinations have a row
-   Interpolating across any missing values within a species time-series
-   Holding the final value constant where a ts ends before 2024

```{r sm_wbi}
data1 <- read.csv("/data/tidy_data/d4a_spp_ts_engto24_Smoothed_unsmVP_120326.csv")

#convert index2 back to measurement scale so tidying rqd and tidy_d functions work properly
data24 <- data1 %>% 
  mutate(index = exp(index2),
         common_name = NA) %>% 
  rename(scientific_name = species)

#run tidying report
tidy_rpt24 <- tidying_rqd(data24)

#expand grid to add missing species year combinations, interpolate missing values within time-series and hold the final value constant where time-series end before the final year. There will be no zeros in the dataset at this stage
data24_geom <- tidy_d(data = data24, tidy_rpt = tidy_rpt24 , exp.g = "Y", interp = "Y", exp_tail = "Y") %>% 
  rescale(ind_col = index, Index_root = 1) %>% #rescale index
  group_by(scientific_name) %>% 
  fill(group, .direction = "updown") %>% 
  fill(survey, .direction ="updown")
  # add group and survey info into new extra cols
    
#calculate geometric mean all and by group
all_geo <- indFun1(data24_geom, ind_col = index, grp_col = scientific_name) %>% 
  mutate(group = "all")

grp_geo <- ddply(data24_geom, .(group), indFun1, ind_col = index, grp_col = scientific_name) %>% 
  full_join(all_geo) %>% 
  rename(unsmoothedMSI = comp_index) %>% 
  mutate(unsmoothedMSI = unsmoothedMSI*100)

#merge with freeman model outputs
df <- df %>% 
  merge(grp_geo, by.x=c("Year", "group"), by.y=c("year", "group"), all = TRUE)
```

If we want to compare to an actual unsmoothed version, we would need to
use the unsmoothed time-series and then process and calculate the
geometric mean using 'IndFun'

Read in unsmoothed time-series and do the same processing as above to
get dataset ready to calculate the geometric mean

```{r unsm_wbi}
data_un <- read.csv("/data/tidy_data/d4_spp_timeseries_engto24_120326.csv")

data_un24 <- data_un %>% 
  rename(index_log = index,
    scientific_name = scientific_name_data) %>% 
  rename(index = index_m)

#run tidying report
tidy_rpt_un24 <- tidying_rqd(data_un24)

#expand grid to add missing species year combinations, interpolate missing values within time-series and hold the final value constant where time-series end before the final year. There will be no zeros in the dataset at this stage
data_un24_geom <- tidy_d(data = data_un24, tidy_rpt = tidy_rpt_un24 , 
                         exp.g = "Y", interp = "Y", exp_tail = "Y") %>% 
  rescale(ind_col = index, Index_root = 1) %>% #rescale index
  group_by(scientific_name) %>% 
  fill(group, .direction = "updown") %>% 
  fill(survey, .direction ="updown")
  # add group and survey info into new extra cols
    
#calculate geometric mean all and by group
all_geo_un <- indFun1(data_un24_geom, ind_col = index, grp_col = scientific_name) %>% 
  mutate(group = "all")

grp_geo_un <- ddply(data_un24_geom, .(group), indFun1, ind_col = index, 
                    grp_col = scientific_name) %>% 
  full_join(all_geo_un) %>% 
  rename(unsmoothedMSI = comp_index) %>% 
  mutate(unsmoothedMSI = unsmoothedMSI*100)
```

plot geomean based on unsmoothed time-series

```{r, fig.width=12, fig.height=6}
ggplot(grp_geo_un, aes(x = year, y = unsmoothedMSI)) +
  geom_line()+
  geom_point()+
  facet_wrap(~group, scales = "free_y")+
  theme_bw()
```

# Plot indicator

For simplicity we will plot only the smoothed M index.

## Headine indicator

```{r plot, fig.width=12, fig.height=6}
gp0 <- ggplot(data = subset(df, group=="all"), aes(x=Year)) +
  geom_ribbon(aes(ymin=lowerCI.M, ymax=upperCI.M), alpha = 0.1) +
  geom_line(aes(y=Index.M)) 
  geom_point(aes(y=unsmoothedMSI), col="red") +  
  facet_wrap(~ypk, nrow=2) +
  geom_hline(yintercept = 100) +
  ylab("Index (baseline = 100)") +
  theme_bw()
gp0
```

Each panel here represents a different degree of smoothing. In all
cases, smoothing on the log scale means that the long term trend is for
a larger decline (than smoothing on the measurement scale)

You can see clearly that, in the case with one knot per three years, the
smoothed and unsmoothed lines are virtually identical. The impact of the
smoothing becomes progressively clear as the smoothing window (number of
years per knot) increases.

The red dots here are from the WBI method for calculating the geometric
mean. It's the same data plotted in all the panels. We would expect this
to map on very closely to the 1 year per knot example. 

## By taxon

Plot separately for each taxon.

```{r plot1, fig.width=12, fig.height=12}
gp1 <- ggplot(data = df, aes(x=Year)) +
  geom_ribbon(aes(ymin=lowerCI.M, ymax=upperCI.M), alpha = 0.1) +
  geom_line(aes(y=Index.M)) +
  geom_point(aes(y=unsmoothedMSI), col="red", size = .4) +  
  facet_grid(group~ypk, scales = "free_y") +
  ylab("Index (baseline = 100)") +
  theme_bw()

gp1  %+% subset(df, !group %in% c("all", "all_five_yr", "all_ten_yr")) + 
geom_hline(yintercept = 100)
```

You can really see the effect of changing the number of knots on the
moth trend.

Zoom into the period since 2000 (including the "all" groups now)

```{r, fig.width=12, fig.height=10}
minYr <- 2000
gp1  %+% subset(df, !group %in% c("all_five_yr", "all_ten_yr", "all")) + 
  xlim(c(minYr, 2025))
```

In these plots, the uncertainty is expressed relative to the baseline
(1970 for some groups). I could re-calculate the uncertainty ribbons so
that they're measuring change relative to this new baseline.

plot re-baselined all species 3ypk

```{r}
ggplot(data = subset(df, group %in% c("all", "all_five_yr", "all_ten_yr") & ypk == 3), aes(x=Year)) +
  geom_ribbon(aes(ymin = lowerCI.M, ymax = upperCI.M, fill = group), alpha = 0.2 ) +
  geom_line(aes(y = Index.M, col = group)) +
  geom_hline(yintercept = 100) +
  ylab("Index (baseline = 100)") +
  scale_x_continuous(breaks = breaks_pretty()) +
  theme_bw()
```

rescale to final year to make comparison easier and look post 2010

```{r rebase}
rb <- df %>% 
  filter(group == "all" & ypk == 3| group ==  "all_five_yr"| group == "all_ten_yr" ) %>% 
  group_by(group) %>% 
  mutate(rel_final = Index.M/Index.M[Year == 2024])

ggplot(rb, aes(x=Year)) +
  geom_line(aes(y = rel_final, col = group)) +
  ylab("Index (final year = 1)") +
  scale_y_continuous( limits = c(.75, 2)) +
  theme_bw() +
  xlim(2010, 2025)
```

Differences between the different numbers of knots are not especially
clear.

```{r plotM2, fig.width=12, echo=FALSE}
ggplot(data = subset(df), aes(x=Year)) +
  geom_ribbon(aes(ymin=lowerCI.M, ymax=upperCI.M, fill=ypk), alpha = 0.2) +
  geom_line(aes(y=Index.M, col=ypk)) +
  facet_wrap(~group, ncol=3, scales = "free") +
  geom_hline(yintercept = 100) +
  ylab("Index (baseline = 100)") +
  scale_x_continuous(breaks = breaks_pretty()) +
  theme_bw()
```

For most groups, the choice of smoothing makes very little difference.
There is an observable effect for plants, moths and butterflies.

Now overlay the different Mprime (unsmoothed indicator) lines

```{r plotM3, fig.width=12, echo=FALSE}
ggplot(data = subset(df), aes(x=Year)) +
  geom_line(aes(y=Index.Mprime, col=ypk)) +
  facet_wrap(~group, ncol=3, scales = "free") +
  geom_hline(yintercept = 100) +
  ylab("Index (baseline = 100)") +
  scale_x_continuous(breaks = breaks_pretty()) +
  theme_bw()
```

This shows that the "unsmoothed" line (M.prime) is almost completely
insensitive to the number of knots. This is particularly true for the
taxon-specific trends, but less true of the all-species plot and plants.

## Official plot

Defra has decided that the preferred option is to present the
`ypk = c(10, 3)` lines on the same plot, with `ypk = 1` as points
representing the unsmoothed trend.

```{r}
prodplotAll <- ggplot(data = subset(df, group %in% c("all") & ypk %in% c(3, 10)), 
                      aes(x=Year)) +
  geom_ribbon(aes(ymin=lowerCI.M, ymax=upperCI.M, fill=ypk), alpha = 0.2) +
  geom_line(aes(y=Index.M, col=ypk)) +
  geom_point(data = subset(df, group %in% c("all") & ypk == 1), 
             aes(x=Year, y=Index.M), col = "black") +
  geom_hline(yintercept = 100) +
  ylab("Index (baseline = 100)") +
  scale_x_continuous(breaks = breaks_pretty()) +
  theme_bw()
prodplotAll
```

The unsmoothed (1 knot per year) line is almost identical to the 3
years-per-knot line.

Now for each taxonomic group

```{r, fig.width = 12}
prodplotTx <- ggplot(data = subset(df, !group %in% c("all", "all_five_yr", "all_ten_yr") & ypk %in% c(3, 10)), 
                      aes(x=Year)) +
  geom_ribbon(aes(ymin=lowerCI.M, ymax=upperCI.M, fill=ypk), alpha = 0.2) +
  geom_line(aes(y=Index.M, col=ypk)) +
  geom_point(data = subset(df, !group %in% c("all") & ypk == 1), 
             aes(x=Year, y=Index.M), col = "black", size = .5) +
  geom_hline(yintercept = 100) +
  ylab("Index (baseline = 100)") +
  scale_x_continuous(breaks = breaks_pretty()) +
  theme_bw()

prodplotTx + facet_wrap(~group, nrow=4, scales = "free")
```

We can see that the unsmoothed index is almost identical to the 3 years
per knot line.

NB several time-series don't quite have enough data for 10 years per
knot.

-   Bumblebees have 15 years of data, so we would really need one more
    year for a second knot, according to the rules we set.
-   Freshwater invertebrates have only 12 years of data; vascular plants
    ten.
-   In these cases the number of knots is two. This corresponds to 7.5
    ypk for bumblebees, 6 ypk for FWI and 5 ypk for plants. We need to
    capture this in the indicator documents.

# Change metrics

Now summarise the long-term, short-term and instantaneous change for
each version of the index.

## Long term

The long term trend is the same, regardless of the number of knots.
Let's report these as % change since the baseline.

```{r}
sf <- 3
LongTerm <- subset(df, ypk == 10, select = !grepl("prime", names(df))) %>% 
  group_by(group) %>%
  filter(Year == max(Year)) %>%
  mutate(Index.M = signif(Index.M - 100, sf),
         lowerCI.M = signif(lowerCI.M - 100, sf),
         upperCI.M = signif(upperCI.M - 100, sf)) %>%
  dplyr::rename(BestEst = Index.M,
                lowerCI = lowerCI.M,
                upperCI = upperCI.M)

LongTerm[,c(6,5,2:4)]
```

The lower and upper CI represent the 95% credible intervals. This means
we are 95% confident that the true value lies within this range.

## Short Term assessment

We would need to go back to the `output` object in order to get the CI.
For now, then, I'll just report the best estimate of change over the
past 5 years (i.e. 2017-2022 for the most)

```{r}
TimeboundChange <- function(x, metric="ind", t=5){
  temp <- subset(df, group == x["group"][[1]] & ypk == as.numeric(x["ypk"][[1]])) 
  Y1 <- max(temp$Year, na.rm=TRUE)
  Y0 <- Y1 - t
  temp2 <- subset(temp, Year %in% c(Y0, Y1))
  stat <- diff(temp2$Index.M)
  if(metric == "pc"){ # convert to a percentage
    stat <- 100 * stat / temp2[temp2$Year==Y0,"Index.M"]
  }
  round(stat,1)
}

tempch <- apply(pars, 1, TimeboundChange, metric="ind", t=5)
ShortTerm1 <- cbind(pars[c(1,2,5)], change=tempch)
(ShortTerm1 <- acast(ShortTerm1, group~ypk, value.var="change"))
```

Each column represents a different degree of smoothing (number of knots
per year). Note that, when pre-smoothing happens on the log scale, there
is very little variation in the 5 year trend. When smoothed on the
measurement scale, the magnitude of change varies by 100% for moths and
butterflies.

Values in this table are relative to the baseline conditions in the
first year. Let's repeat these calculations but calculate change as a
percentage of the value five years ago.

```{r}
change <- apply(pars, 1, TimeboundChange, metric="pc", t=5)
ShortTerm1 <- cbind(pars[c(1,2,5)], change=change)
(ShortTerm1 <- acast(ShortTerm1, group~ypk, value.var="change"))
```

## Final year change

Finally, let's look at the value of the first derivative in the final
year. Again, we'll express this as a percentage of the previous year's
value.

```{r}
finalYrChange <- function(mod){
    modsum <- attr(mod, "model")$summary
    logLam <- modsum[grepl("logLambda", dimnames(modsum)[[1]]),]
    ll <- logLam[nrow(logLam),c(1,3,7,11)]
    ll[1:3] <- 100 * (exp(ll[1:3]) - 1)
    signif(ll, 2)
}

# log scale pre-smoothing
FinalYr <- t(sapply(output, finalYrChange))
(FinalYr <- cbind(pars[,c(1,2,3)], FinalYr))
```

The `f` column indicates the proportion of the posterior distribution
that is the same sign as the mean, i.e. our confidence in the direction
of the trend. Values close to 1 indicate high confidence; values close
to 0.5 (the minimum) indicate low confidence in a change. Note that, in
most cases, our estimates are quite confident (i.e. `f > 0.8`), meaning
that we reject the "no change" category.

### Summary of final year

In general there is both quantitative and qualitative agreement between
the different degrees of smoothing: 

- Birds are declining by 0.72-1.6%pa 
- Bumblebees are declining at 7.5-8.1% pa 
- Butterflies are declining by 16-17% pa 
- freshwater invertebrates are declining by 0.52 to increasing by 0.24% pa 
- Mammals are increasing at 1.2to1.3% pa 
- Moths are declining by 9.1 - 12% pa 
- Fish are declining at 1.2 to 2.6% pa 
- Plants are increasing by 1.8 to 2.6 % pa

-   For "all" there is a decline of 4.7 to 5.6%pa for three of the
    smoothed options.

# Species-level trends

We first need to identify the two "all species" runs

```{r}
#we need to filter out the re-based model outputs
output     <- output[-grep("_yr", names(output))]
#need to remove re-based models for this to work properly 
modelfiles <- modelfiles[-grep("_yr", modelfiles)]
```

```{r}
getSpTrends <- function(name, n = 0){
  # `n` is an optional variable determining the number of years over which to estimate the trend
  # if zero then the full series is used. If an integer above zero then the final `n` years of the series is used 
  mod <- attr(output[[grep(name, modelfiles)]], "model")
  spg <- mod$mean$spgrowth
  estimate <- mod$data$estimate # the raw data. Has one more column than spg

  if(n > 0){
    # short term assessment. Restrict spg to the last n years
    N <- dim(spg)[2]
    spg <- spg[,(N-n):N]
    estimate <- estimate[,(N-n):N]
  } else {
    estimate <- estimate[,-1]
  }
  
  spMGR <- sapply(1:nrow(spg), function(i){
    rates <- spg[i,!is.na(estimate[i,])]
    mean(rates[-1]) # ignore first value - this is the first year with data so the GR is from other species
  })
  
  names(spMGR) <- dimnames(estimate)[[1]]
  100*(exp(spMGR)-1)
}

spTrends <- lapply(c(0,5), function(n){
  x<-as.data.frame(do.call(cbind, lapply(list("all_3", "all_10"), getSpTrends, n=n)))
  names(x) <- paste0(c("ypk3", "ypk10"), "_",n)
  x
})
spTrends <- do.call(cbind, spTrends)
head(spTrends)
```

The `ypk3_0` column is the long term trend when smoothed with three
knots per year. The `ypk10_5` column is the short term trend (last 5
years) when smoothed with ten knots per year. etc etc. All trends are
measured as the mean across years, converted to a percentage
increase/decrease in abundance

Take a look at the overall distribution

```{r}
ggplot(data=melt(spTrends)) + 
  geom_freqpoly(aes(value, col=variable), bins = 25) +
  xlab("species trend (%/yr)") +
  theme_bw()
```

Some very large values! Let's look at the quantiles.

```{r}
pc <-  c(1,5,10,25)
signif(apply(spTrends, 2, quantile, c(pc/100,0.5,sort(1-pc/100)), na.rm=T),2)
```

We need to include the `na.rm=T` because there is a small number of
species with no data in the last 5 years

This clearly shows that most species trends are within the range of what
we would expect. The short-term trends are absolutely larger than the
long-term: again this is what we expect (short term trends are subject
to noisy fluctuation).

For the long-term trends:

-   half of all species are in the range [-2.3 to +1.7 pa]
-   80% of species are within the range [ -4.8 to +4.3 pa]
-   90% of species are in the range [-6.7 to +6.6 pa]
-   99% of species are in the range [-11 to +13 pa]
-   There is a small minority of species with extreme trends outside
    this range

## Extreme species trends

Take a look at the raw data for the most steeply declining species

```{r}
losers <- subset(spTrends, ypk3_0 < -11)
losers
ggplot(data = subset(data1, species %in% rownames(losers)),
       aes(x=year, y=index2, col=species)) +
    geom_line() +
  theme_bw()
```

-   Moths
    -   Abraxas sylavata - clouded magpie zeros in many recent years
    -   Orthonama vittata- oblique carpet zeros in some recent years
-   Bumblebee - Bombus campestris
-   Fw invert
    -   Halipius confinis zero in 2024
    -   Lmnephilus extricatus
    -   Notonecta viridis
    -   Rhyacophilia fasciata
-   Plants
    -   Dioscorea communis aka tamus communis
    -   Hordeum secalinum
-   Fish
    -   Osmerus eperianus smelt

And the steepest increases

```{r}
winners <- subset(spTrends, ypk3_0 > +13)
winners
ggplot(data = subset(data1, species %in% rownames(winners)),
       aes(x=year, y=index2, col=species)) +
    geom_line() +
  theme_bw()
```

-   Fish
    -   Ammodytes tobianus - lesser sandeel, sparse data in recent
        years, smoothed trend looks odd
-   Other species trends seem to reflect underlying data
    -   Clupea harengus - herring
    -   Liza ramada - thin lipped grey mullet
    -   Cobitis taenia - spined loach
    -   Pomatoschistus microps - common goby
-   Bumblebees: Bombus lucorum/terrestris
-   Moth: Mythiman ablipuncta - white-point
-   FW inv: Dryops spp., Nigrobaeis spp, Potamanthus luteus
-   Bird: Milvus milvus - Red Kite (reinforcement), Platalea
    leucorodia - spoonbill (coloniser)
-   Butterfly: Maculinea arion - Large Blue, reintroduced, rapid
    recovery

## Categories of change

Get the group identity and summarise into categories and report the
values for each group

```{r}
spTrends$group <- data1$group[match(rownames(spTrends), data1$species)]
spTrends$survey <- data1$survey[match(rownames(spTrends), data1$species)]

spTrends$longTrendCat <- cut(spTrends$ypk3_0, 
                         breaks = c(-Inf, -2.73, -1.14, +1.16, +2.81, Inf),
                         labels = c("Strong decrease", "Weak decrease", "Little change", "Weak increase", "Strong increase"))
summary(spTrends$longTrendCat)

spTrends$shortTrendCat <- cut(spTrends$ypk3_5, 
                         breaks = c(-Inf, -2.73, -1.14, +1.16, +2.81, Inf),
                         labels = c("Strong decrease", "Weak decrease", "Little change", "Weak increase", "Strong increase"))
summary(spTrends$shortTrendCat)

with(spTrends, table(group, longTrendCat))
with(spTrends, table(group, shortTrendCat))
```

```{r, eval=FALSE}
write.csv(spTrends, file = file.path(modeldir,"spTrends_allspp_to2024_130326.csv")) # unsm VP cat chg
```

Summarise species trends - based on categorisation of change from knots
per 3 year model

```{r}
spTrend_grp <- spTrends %>% 
  select(group, longTrendCat, shortTrendCat) %>% 
  melt(id.vars = "group") %>% 
  group_by(group, variable, value) %>% 
  summarise(n_spp = n()) %>% 
  group_by(group, variable) %>% 
  mutate(pct_spp = n_spp/sum(n_spp[!is.na(value)]))
         
spTrend_sum <- spTrends %>% 
  select(group, longTrendCat, shortTrendCat) %>% 
  melt(id.vars = "group") %>% 
  group_by(variable, value) %>% 
  summarise(n_spp = n()) %>% 
  group_by(variable) %>% 
  mutate(pct_spp = n_spp/sum(n_spp[!is.na(value)]),
         group = "all") %>% 
  full_join(spTrend_grp) %>% 
  rename(period = variable, trend_category = value) %>% 
  mutate(trend_category = 
           factor(trend_category,
                  levels = c("Strong decrease", "Weak decrease", "Little change",
                             "Weak increase", "Strong increase"))) %>% 
  arrange(group, period, trend_category)
```

```{r, eval=FALSE}
write.csv(spTrend_sum, file = file.path(modeldir, "spTrends_allspp_to2024_summary130326.csv"), row.names = FALSE)
```
