---
title: "run Freeman By Group"
author: "Nick Isaac (original)"
date: "2024-02-28 (original)"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

We're going to run the Freeman method

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(devtools)
library(BRCindicators)
library(jagsUI)
library(ggplot2)
library(reshape2)
```

Load in the data.

```{r readData}
datadir <- "/data/tidy_data/2025" 

# to use unsmoothed VP time-series
data <- read.csv(file = file.path(datadir,"d4a_spp_ts_engto23_Smoothed_unsmVP.csv")) 
```

Use smoothing on the log scale (`index2`).

```{r}
logSmoothing <- TRUE
if(logSmoothing){
  data$index <- data$index2
} else {
  data$index <- data$index1
}
```

As there is no fish data for 2020, we need to add an empty year in 2020. If we 
don't do this, the model will incorrectly believe that 2022 is actually 2021. 
Just need to do this for a single species, so we use Sprattus sprattus.

```{r}
data <- rbind(data[,c("species", "year", "group", "index")], 
              data.frame(species = "Sprattus sprattus", 
                               year = 2020,
                               group = "fish",
                               index = NA))
```

First, define a function to run the models for a particular set of data

```{r runFreeman, include=FALSE, echo=TRUE}
runFreeman <- function(gr, ypk, save = TRUE, runShort = TRUE, nit=5000, spTrends = FALSE){
                       
                       if(gr == "all") {
                         inddat <- data[,c("species","year", "index")]
                       } else {
                         inddat <- subset(data, group==gr, select = c("species","year", "index"))
                       }

                       n.yr    <- length(unique(inddat$year))
                       n.knots <- round(n.yr / ypk)
                       print(paste0(gr, ": ", ypk, " years per knot. With ",n.yr, " years of data = ", n.knots," knots overall"))

                       if(runShort & n.knots == 1) {
                         n.knots <- 2
                         print("running with two knots")
                       }
                       
                       if(n.knots > 1){
                   
                         mod <- bma(data = inddat, 
                                 num.knots = n.knots,
                                 parallel = TRUE,
                                 plot = FALSE,
                                 n.iter = nit,
                                 save.sppars = spTrends)
                       
                         md <- data.frame(group = gr,
                                  ypk = ypk,
                                  n.knots = n.knots,
                                  n.it = nit)
                         attr(mod, "metadata") <- md

                        if(save){
                         letter <- ifelse(logSmoothing, "L","M")
                          #saveRDS(mod, file=paste0("results/", gr, "_",
                          saveRDS(mod, file=paste0("results23/all_species/", gr, "_",
                                                 ypk,"ypk_",letter,"_",
                                                  nit/1000,"kit_",
                                                  format(Sys.Date(),"%y%m%d"),".rds"))
                         return(NULL)
                       } else {
                         return(mod)
                       }
                       } else {
                         print("only one knot: skipping")
                       }
                       
}
```

Define some simple parameters for the model runs.

```{r setPars}
taxa <- unique(data$group)
# set the number of years per knot
ypk <- c(10, 3)
```

Run the vascular plants using the unsmoothed species time-series.

First, we'll run this for all the groups except moths

```{r, echo=FALSE}
p <- expand.grid(ypk = ypk, gr = setdiff(taxa, c("moths", "fish")))
system.time({
    output_group <- mapply(ypk = p$ypk, gr = p$gr, runFreeman, SIMPLIFY = FALSE, save = TRUE)
})
```

Runtime varies with number of species and length of time-series.

Now for the moths

```{r, echo=FALSE}
system.time({
    output_group <- sapply(ypk, runFreeman, gr = "moths", save = TRUE)
})
```

Each run of the moths takes just over 30 minutes

Now the all the species MSI. For this we want to save the species
trends.

```{r, echo=FALSE}
system.time({
    output_group <- sapply(ypk, runFreeman, gr = "all", save = TRUE, spTrends = TRUE)
})
```

Just over an hour each (a bit slower if saving species parameters)

Finally, we'll rerun each group with one knot for every year of data.

```{r}
system.time({
    output_group <- sapply(c(taxa, "all"), runFreeman, ypk = 1, save = TRUE, spTrends = FALSE)
})
```

