---
title: "English abundance index"
author: "Nick Isaac, original"
date: "2024-03-22, original date"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
---

We have run the Freeman method for multiple groups on a dataset of pre-smoothed index values (on the log scale).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(coda)
library(ggplot2)
library(reshape2)
library(plyr)
library(dplyr)
library(tidyr)
library(scales)
library(BRCindicators)
```


# Read in the model files
```{r readData, include = FALSE}
modeldir <- "results23/all_species" # Freeman output model location
```

This dataset contains pre-smoothed the species trends. We have checked the warnings and found there is no real problem here. We have used the smoothed trends except for:
- Bats, which were pre-smoothed.
- for two species with only two years of data
- where the smoother failed (both species have only three years of data)
- A small number of cases (n=25 species:year combinations) where the fitted count is negative
- NB in previous runs we did not pre-smooth the vascular plant trends. This has changed.

## Get the models
```{r setPars}
modelfiles <- list.files(modeldir, recursive = TRUE, patt = ".rds")
head(modelfiles)
```

## load the models
```{r}
output <- lapply(modelfiles, function(modfilename){
 readRDS(file.path(modeldir, modfilename)) 
})
```

## Get the metadata
Extract the metadata from all the model runs into a dataframe
```{r}
pars <- lapply(output, function(mod){
  md <- attr(mod, "metadata")
})
pars <- do.call(rbind, pars)
```

## Explore and combine
Has it converged?
```{r}
MCMCStats <- lapply(output, function(msi) {
  dat <- attr(msi, "model")$summary
  dat[!grepl("^M", dimnames(dat)[[1]]),8]
})
pars$Rhat3q <- sapply(MCMCStats, quantile, 0.75, na.rm=TRUE)
```


Combine them all together
```{r}
df <- do.call(rbind, output)
attr(df, "model") <- NULL
attr(df, "metadata") <- NULL
```


Now coerce the output into a useful format:
```{r Manipulate}
pars$nyr <- c(sapply(output, nrow))

df$group <- melt(mapply(rep, pars$group, pars$nyr))$value
df$ypk <- melt(mapply(rep, pars$ypk, pars$nyr))$value
df$n.knots <- melt(mapply(rep, pars$n.knots, pars$nyr))$value
df$ypk <- factor(df$ypk)
```

Save the data to send to Defra who produce the plots from the csv files
```{r}
write.csv(df,  file = file.path(modeldir, "MSI|group+_presmooth_allspp_to2023.csv"))
write.csv(pars, file = file.path(modeldir, "SmoothingParameters_allspp_to2023.csv"))
```

# Species-level trends

We first need to identify the two "all species" runs
```{r}
getSpTrends <- function(name, n = 0){
  # `n` is an optional variable determing the number of years over which to estimate the trend
  # if zero then the full series is used. If an integer above zero then the final `n` years of the series is used 
  mod <- attr(output[[grep(name, modelfiles)]], "model")
  spg <- mod$mean$spgrowth
  estimate <- mod$data$estimate # the raw data. Has one more column than spg

  if(n > 0){
    # short term assessment. Restrict spg to the last n years
    N <- dim(spg)[2]
    spg <- spg[,(N-n):N]
    estimate <- estimate[,(N-n):N]
  } else {
    estimate <- estimate[,-1]
  }
  
  spMGR <- sapply(1:nrow(spg), function(i){
    rates <- spg[i,!is.na(estimate[i,])]
    mean(rates[-1]) # ignore first value - this is the first year with data so the GR is from other species
  })
  
  names(spMGR) <- dimnames(estimate)[[1]]
  100*(exp(spMGR)-1)
}

spTrends <- lapply(c(0,5), function(n){
  x<-as.data.frame(do.call(cbind, lapply(list("all_3", "all_10"), getSpTrends, n=n)))
  names(x) <- paste0(c("ypk3", "ypk10"), "_",n)
  x
})
spTrends <- do.call(cbind, spTrends)
head(spTrends)
```

The `ypk3_0` column is the long term trend when smoothed with three knots per year.
The `ypk10_5` column is the short term trend (last 5 years) when smoothed with ten knots per year.
etc etc.

All trends are measured as the mean across years, converted to a percentage increase/decrease in abundance

Take a look at the overall distribution
```{r}
ggplot(data=melt(spTrends)) + 
  geom_freqpoly(aes(value, col=variable), bins = 25) +
  xlab("species trend (%/yr)") +
  xlim(c(-30,30))+
  theme_bw()
```

## Categories of change

Get the group identity and summarise into categories and report the values for each group
```{r}
data1 <- read.csv("../../data/d4a_spp_ts_engto23_Smoothed_unsmVP.csv")

spTrends$group <- data1$group[match(rownames(spTrends), data1$species)]
spTrends$survey <- data1$survey[match(rownames(spTrends), data1$species)]

spTrends$longTrendCat <- cut(spTrends$ypk3_0, 
                         breaks = c(-Inf, -2.73, -1.14, +1.16, +2.81, Inf),
                         labels = c("Strong decrease", "Weak decrease", "Little change", "Weak increase", "Strong increase"))
summary(spTrends$longTrendCat)

spTrends$shortTrendCat <- cut(spTrends$ypk3_5, 
                         breaks = c(-Inf, -2.73, -1.14, +1.16, +2.81, Inf),
                         labels = c("Strong decrease", "Weak decrease", "Little change", "Weak increase", "Strong increase"))
summary(spTrends$shortTrendCat)

with(spTrends, table(group, longTrendCat))
with(spTrends, table(group, shortTrendCat))


write.csv(spTrends, file = file.path(modeldir,"spTrends_allspp_to2023.csv")) # unsm VP cat chg
```




